Polynomial Time and Dependent Types

### 2023-12-14 paper/paper.tex
[ paper ] Conference Publishing's demands

\subsection{Contributions and Content}

@@

\subsection{The Cons-free System}
\label{sec:cons-free-intro}

@@

\subsubsection{Natural-number Usages}

@@

\subsection{Non-iterable Data Types}
\label{sec:noniter-qtt}

@@

\subsection{Cons-free Natural Numbers and their Recursor}
\label{sec:cons-free-qtt}

@@

\subsection{LFPL-style Diamonds, Natural Numbers, and a Recursor that Gives Back}
\label{sec:lfpl-qtt}

@@

\subsection{Polytime-based Complexity Classes}
\label{sec:classes}

@@

\subsubsection{Bounded-error Probabilistic Polynomial Time}

@@

In this section and the next, we establish the polytime soundness of
our extensions of QTT by adapting a realisability method due to
\citet{dallago11realisability}. This approach is based on a three way
coupling between abstract mathematical elements (the \emph{what}),
values from a machine model (the \emph{how}), and resource potentials
(the \emph{fuel}). Each type in the system is defined as a three way
relation between these elements. The set of abstract elements depends
on the type being interpreted (e.g., types of natural numbers will be
defined in terms of the set $\mathbb{N}$). The machine model is fixed
across all types. We describe the particular machine model we use for
this paper in \autoref{sec:machine-model}. Potentials are arranged
into \emph{resource monoids} that we define in
\autoref{sec:resource-monoids}. Unlike \citet{dallago11realisability},
we explicitly construct realisers for inductive datatypes (both
iterable and non-iterable) instead of relying on second-order
polymorphic encodings and special $\oc$-style modalities. These
explicit constructions are essential to construct models of our systems.

@@

Costed evaluation of expressions in environments is defined by a
big-step operational semantics $E, \eta \Downarrow_k V$, where $k$ is
the number of steps. For simplicity, all operations cost $1$ unit,
though this could be generalised to allow for different operations to
have different costs. We use $\eta[i]$ to access the $i$th variable in
the environment, counting from the right. The evaluation rules are
mostly as one would expect, except for the application rule which
includes a self reference to the closure being invoked, in order to
allow recursive definitions.

\paragraph{Agda Formalisation} The machine model is defined in the
Agda module \AgdaModule{MachineModel}. We use an intrinsically
well-scoped syntax, which ensures that all variable look up operations
are always well defined.

@@

\subsubsection{Resource sub-monoids}

@@

\subsubsection{Non-iterable Data Types}
\label{sec:noniter-model}

>>>>

\paragraph{Contributions and Content}

@@

\subsection{The Cons-Free System}
\label{sec:cons-free-intro}

@@

\subsubsection{Natural-Number Usages}

@@

\subsection{Non-Iterable Data Types}
\label{sec:noniter-qtt}

@@

\subsection{Cons-Free Natural Numbers and their Recursor}
\label{sec:cons-free-qtt}

@@

\subsection{LFPL-Style Diamonds, Natural Numbers, and a Recursor that Gives Back}
\label{sec:lfpl-qtt}

@@

\subsection{Polytime-Based Complexity Classes}
\label{sec:classes}

@@

\subsubsection{Bounded-Error Probabilistic Polynomial Time}

@@

In this section and the next, we establish the polytime soundness of
our extensions of QTT by adapting a realisability method due to
\citet{dallago11realisability}. This approach is based on a three way
coupling between abstract mathematical elements (the \emph{what}),
values from a machine model (the \emph{how}), and resource potentials
(the \emph{fuel}). Each type in the system is defined as a three way
relation between these elements. The set of abstract elements depends
on the type being interpreted (e.g., types of natural numbers will be
defined in terms of the set $\mathbb{N}$). The machine model is fixed
across all types. We describe the particular machine model we use for
this paper in \autoref{sec:machine-model}. Potentials are arranged
into \emph{resource monoids} that we define in
\autoref{sec:resource-monoids}. Unlike \citet{dallago11realisability},
we explicitly construct realisers for inductive datatypes (both
iterable and non-iterable) instead of relying on second-order
polymorphic encodings and special $\oc$-style modalities. These
explicit constructions are essential for constructing models of our systems.

@@

Costed evaluation of expressions in environments is defined by a
big-step operational semantics $E, \eta \Downarrow_k V$, where $k$ is
the number of steps. For simplicity, all operations cost $1$ unit,
though this could be generalised. We use $\eta[i]$ to access the $i$th
variable in the environment, counting from the right. The evaluation
rules are mostly as one would expect, except that the application rule
includes a self reference to the closure being invoked in order to
allow recursive definitions.

\paragraph{Agda Formalisation} The machine model is defined in the
Agda module \AgdaModule{MachineModel}. We use an intrinsically
well-scoped syntax, which ensures that all variable accesses are well
defined.

@@

\subsubsection{Resource Sub-Monoids}

@@

\subsubsection{Non-Iterable Data Types}
\label{sec:noniter-model}
